Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex

%% Uncomment below to include a teaser figure.
\teaser{
  \centering
  \begin{tabular}{cc}
  \includegraphics[height=0.22\textheight]{EmailSentRecdSept11toNov23} &
  \includegraphics[height=0.22\textheight]{InnerCircleSept11ToNov23} \\
  &\\
  {\small (a) Missing block of email: October 30 - November 9} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
  \end{tabular}
  \caption{Metadata: Time, date, redaction, correspondents, volume, source type.  Filtered: Sept. 11 to Nov. 23, 2012.  
  }

@@

In this paper, we describe an implementation of the missing visual analytic service and illustrate it on Secretary Clinton's e-mails.  The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in the emails.

The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}. The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}. Within this section, Subsection \ref{sect:Displays:circle} describes the spoked network plot, Subsection \ref{sect:Displays:volume} addresses the email volume time series plot, Subsection \ref{sect:Displays:times} discusses the scatterplot of daily sending times, Subsection \ref{sect:Displays:FOIA} outlines the exemption code barplot, and Subsection \ref{sect:Displays:TFIDF} explains the content tf-idf and frequency information. Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}. Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Subsection \ref{sect:Analysis:Timeline}.

@@

Where possible, the email addresses have also been parsed so that they may be classified into one of four categories: those which are \texttt{.gov},  those which are \texttt{.mil},  those which are identifiable as coming from a domain that is neither  \texttt{.gov} nor  \texttt{.mil}, and those whose domain was not identifiable from the data.
\subsection{Redaction information}
\label{sect:data:redact}
Emails that are redacted are marked as such by the string ``RELEASE IN PART'' and by the presence of one or more of nine FOIA exemption codes B1, B2, \ldots, B9  marking the place where text is missing (redacted).  This provides two further pieces of quasi-metadata (since it is not actual email content) that can be used in analysis.
\subsection{Content information}
\label{sect:data:contentinfo}
The entire (redacted) content of each email is available on the State Department and Wikileaks sites and the user is encouraged to view it there (the pdf forms are more informative, especially in appreciating the extent of the redactions).  To provide some coarse statistical summaries of the content, all word tokens are extracted and partially processed to reduce their number.  For example,  ``stopwords'', as identified by the\texttt{stopwords} function under either the "en" and "SMART" settings, or a set of custom stopwords, were removed. The remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.   While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some insight as to the topics covered.

@@

\subsection{Inner circle}
\label{sect:Displays:circle} 
This display takes up to the 20 most frequent correspondents in the selected emails and arranges their email addresses  at the ends of  equiangular spokes around the hub that is Secretary Clinton's email.  The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).  Every account that is identifiably not a government sponsored account is coloured red.  Those that are identifiably government are coloured either blue (for \texttt{.gov}) or black (for \texttt{.mil}, if any appear).  Those which can not be determined as either government or not, are coloured orange.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
The figure shows the top 20 (or ``Inner Circle'')  hereover all emails in the collection.
Note that Clinton aide Huma Abedin is seen to be using both a \texttt{.gov} account for some email and a \texttt{clintonemail} account for others and so appears as two different colour spokes.  The other identfiably non-government account is that of Sidney Blumenthal.  This is of particular interest, as his closeness to Clinton in providing advice on all kinds of matter has been the source of controversy \cite{BlumenthalControversy} given that he had been refused a State department position and so lacked  State Department clearance.  Wendy Sherman, though appointed by Clinton to  Under Secretary of State for Political Affairs in 2011, her email address could not be identified as either definitively government or definitively non-government and so is shown in orange.  It is important to note, however, that the most likely reason an email is not identifiable is that it has been redacted (likely B6).  While no guarantee that the email is not a government email, it is more likely that it was viewed as personal or private information and so probably not government.

The greater the number of emails between Clinton and the correspondent, the wider is the spoke.  Whenever the difference in volume is great enough the correspondents will also be separated into two groups.  Those with the greatest correspondence will have shorter spokes and hence be visually ``closer'' to Clinton.   As the display shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan. \TODO{Chris, can you be more precise about the algorithm?}
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day. This allows users to view which days have an exceptionally high or low number of classified conversations. Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}

This plot shows highly variable nature of the daily email volumes, with a number of peaks and troughs suggesting regions in which further investigation might be insightful.

\subsection{Email times}
\label{sect:Displays:times}
As shown in Figure~\ref{fig:TimesAll}, this display shows the time stamp of each email separated into its calendar date on the horizontal axis and its (24hr) time of day on the vertical.  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTime}
\caption{Email time patterns: all emails}
\label{fig:TimesAll}
\end{center}
\end{figure}
Each email is a point in the plot and is coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending prevents over-plotting from obscuring patterns.

A few patterns are easily discerned.  For example, the least amount of email appears for a few hours around  20:00 hours, or 8 PM.  Evenings appear to be when email traffic is lightest.  Otherwise, for the most part, it seems fairly uniformly distributed throughout.  Midnight is at the top (and bottom) of the plot, so it would seem that email will occur mainly from midnight until about 4 or 5 PM.   Note also the surprisingly regular horizontal lines that appear across the top of the plot.  This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.  The location of these lines switches exactly whenever daylight savings time switches in North America.

\subsection{FOIA exemptions}
\label{sect:Displays:FOIA}
The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request. The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions. The frequencies of each of these codes within filtered emails are shown in the final display. Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\end{center}

It is clear that redactions made for national security and foreign policy reasons account for only a small portion of redactions made. The most frequent redactions are those made to protect personal privacy, which occur in almost half of all emails, and those made with the vague B5 exemption code, which occur in roughly a quarter of all emails. The other codes appear in very few emails and just over a third of the emails are not redacted at all.

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. \TODO{Chris, check that I have described these two measures correctly}
The top 20 shown here are for all emails in the corpus; these will change depending on the filtering.

The terms give some limited insight into the contents of the email.  As seen here, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.

As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is also challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).   Even so,  the terms nevertheless occasionally turn up something of interest as in the case of ``windrush'',  which turned out to refer to ``Windrush Ventures'' (e.g. see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K. Prime Minister Tony Blair .  ``Windrush'' appears at the end of emails as part of Mr. Blair's electronic signature for his email id ``aclb''  (e.g. see the ``Inner Circle''  of Figure~\ref{fig:teaser}, b).

\section{Filters}
\label{sect:Filters}
There are only three filters but each one affects all data displays.  As each filter is changed, every display redraws itself on the filtered data, immediately in reaction to the change.  In this way, the filters can be used together to focus on particular subsets of the emails or simply to observe how the display patterns change with the filter being applied.
\subsection{Time filtering}
A sliding time window 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{DateSliderImage}
\end{center}
\end{figure}
filters the emails displayed to those whose date lies between the two end points.
The size of the time window is changed by  moving either end, either by mouse click and drag or by selecting an end to move with the arrow keys. The whole time window is moved by selecting the middle bar and either dragging it or moving it with the arrow keys.

This filter is simple but powerful.  All emails are displayed when the range covers all dates.  Moving the end points towards one another allows the user to focus the displays on any particular range, down to as fine as all emails on a single day.  With a fixed range of days, for example a two week period,  dragging the middle bar from left to right will have each display smoothly update over time.
%
%Time filtering within the app can be completed at the level of individual days. The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days. As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width. Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
\subsection{Sent or received}
A drop down menu 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{ToFromImage}
\end{center}
\end{figure}
is used to select those emails which Clinton sent, or received, or either sent or received.  The default is ``All Emails'', which is the same as no filter being applied for sent or received.  Choosing ``From Clinton'' selects for display only those mails sent by Secretary Clinton, choosing ``To  Clinton'' selects only those emails which she received.  This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported in the app with a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{FOIABoxImage}
\end{center}
\end{figure}
Users can add FOIA exemption codes currently not included to the data by clicking the box and selecting a code from the menu which is then displayed below the box. To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu. The box therefore displays all codes currently included in the filter.
\subsection{Auxiliary information}
\TODO{e.g. Clinton's foreign schedule}
A number of cosmetic adjustments using auxiliary information can also be made. Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes. This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic. Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot. This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).

\section{Analysis}
\label{sect:Analysis}

\subsection{Context: A brief timeline on the private email server}
\label{sect:Analysis:Timeline}
While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.  It also introduces some of the principal characters in the whole affair.  More complete and in depth timelines are readily available on the internet (e.g. \cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).

\subsection{Highest peak}
Within the plot of email volume over time, one very large peak stands out.
Historically important,  possibly had an impact on the 2016 US presidential election.

>>>>

%% Uncomment below to include a teaser figure.
\teaser{
  \centering
  \begin{tabular}{cc}
  \includegraphics[height=0.22\textheight]{EmailSentRecdSept11toNov23} &
  \includegraphics[height=0.22\textheight]{InnerCircleSept11ToNov23} \\
  &\\
  {\small (a) An empty email block: October 30 - November 9} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
  \end{tabular}
  \caption{Metadata: Time, date, redaction, correspondents, volume, source type.  Filtered: Sept. 11 to Nov. 23, 2012.  
  }

@@

In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in the emails.

The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}. The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}. Within this section, Subsection \ref{sect:Displays:circle} describes the spoked network plot, Subsection \ref{sect:Displays:volume} addresses the email volume time series plot, Subsection \ref{sect:Displays:times} discusses the scatterplot of daily sending times, Subsection \ref{sect:Displays:FOIA} outlines the exemption code barplot, and Subsection \ref{sect:Displays:TFIDF} explains the content tf-idf and frequency information. Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}. Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Section \ref{sect:Timeline}.

%\section{Background}
%\label{sect:Background}
%
%\subsection{A brief timeline on the private email server}
%\label{sect:Background:Timeline}
\section{A brief timeline on the private email server}
\label{sect:Timeline}
While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.  It also introduces some of the principal characters involved.  More complete and in depth timelines are readily available elsewhere (e.g. \cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).

@@

Where possible, the email addresses have also been parsed so that they may be classified into one of four categories: those which are \texttt{.gov},  those which are \texttt{.mil},  those which are identifiable as coming from a domain that is neither  \texttt{.gov} nor  \texttt{.mil}, and those whose domain was not identifiable from the data.
\subsection{Redaction information}
\label{sect:data:redact}
Emails that are redacted are marked as such by the string ``RELEASE IN PART'' and by the presence of one or more of nine FOIA exemption codes B1, B2, \ldots, B9  marking the place where text is missing (redacted).  This provides two further pieces of quasi-metadata (since it is not actual email content) that can be used in analysis.
\subsection{Content information}
\label{sect:data:contentinfo}
The entire (redacted) content of each email is available on the State Department and Wikileaks sites and the user is encouraged to view it there (the pdf forms are more informative, especially in appreciating the extent of the redactions).  To provide some coarse statistical summaries of the content, all word tokens are extracted and partially processed to reduce their number.  For example,  ``stopwords'', as identified by the \texttt{stopwords} function (with both  \texttt{kind = "en"} and  \texttt{kind = "SMART"}), or as from a set of custom stopwords, were removed.  Remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.   While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some hints about the topics covered.

@@

\subsection{Inner circle}
\label{sect:Displays:circle} 
This display takes the most frequent correspondents (max. 20) in the selected emails and arranges their email addresses  at the ends of  equiangular spokes having Secretary Clinton as hub.  The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).  Every account that is identifiably not a government sponsored account is coloured red.  Those that are identifiably government are coloured either blue (for \texttt{.gov}) or black (for \texttt{.mil}, if any appear).  Those which can not be determined as either government or not, are coloured orange.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided.

The greater the number of emails between correspondents, the wider is their connecting spoke.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.   As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan. \TODO{Chris, can you be more precise about the algorithm?}
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.

Over the four year scale shown in Figure~\ref{fig:VolumeAll}, the display is quite busy.  Even so, a few features stand out.  For example, there is a notable lack of email from the first few months of Clinton's tenure;  there is essentially no email from Clinton at the beginning and very little to her.

One also notices the spikes of email activity.  These can be checked against world events to generate hypotheses about what might be occurring to cause such spikes.  For example, the largest spike occurs on August 21, 2011, the beginning day of the battle for Tripoli and on which it was reported that two of Qaddafi's sons had been captured \cite{battleTripoli}.  It is also the day on which the famous ``tick tock Libya'' email is composed by Jake Sullivan providing a timeline crediting Secretary Clinton with leading the U.S.  policy on Libya ``from start to finish'' \cite{tickTockLibya}.  Much of the email that day is redacted (B5 primarily).   Other peaks (and valleys) of activity could be similarly investigated.

\subsection{Email times}
\label{sect:Displays:times}
Figure~\ref{fig:TimesAll}  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
\caption{Email time patterns: all emails sent from Clinton}
\label{fig:TimesAll}
\end{center}
\end{figure}
shows all emails sent by Clinton over the four year period.  Each email appears as a point determined by the time shown that it was sent.  The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.   Points are coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending is used to minimize the effect of over-plotting.

A few patterns are easily discerned.  For example, the least amount of email appears for a few hours around  20:00 hours, or 8 PM.  Evenings appear to be when email traffic is lightest.  Otherwise, for the most part, it seems fairly uniformly distributed throughout.  Midnight is at the top (and bottom) of the plot, so it would seem that email will occur mainly from midnight until about 4 or 5 PM.

If these are the actual times Secretary Clinton composed and sent email, it suggests that much of that activity occurred in the middle of the night.  To reflect the low email activity period as actual ``downtime'' between ``days'' (now interpreted as separable 24 hour email cycles),  we offer the possibility of choosing 6 PM as the ``end'' of a day rather than midnight.   This places the least active email period as a ``quiet'' boundary between days.   Unfortunately, though it better represents the rhythm of correspondence, it can be confusing since it means that the ``next day''  begins just after 6 PM.

Note also the surprisingly regular horizontal lines that appear across the top of the plot.  This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.  The location of these lines switches exactly whenever daylight savings time switches in North America.

When the received emails are added to the plot, the pattern is essentially the same (though much denser) and the horizontal lines where the server has scheduled something are even clearer.

\subsection{FOIA exemptions}
\label{sect:Displays:FOIA}
U.S. Freedom of Information Act (FOIA) exemption codes B1 through B9 are used in each place in an email where information has been redacted.  For the authoritative definition the act itself should be consulted \cite{FOIA}.  Briefly, they mark each redaction as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memorandums,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.   Each email can contain redactions for any number of exemption codes.

Figure \ref{fig:FOIAcodes}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\caption{FOIA barplot: Exemption codes from all email}
\label{fig:FOIAcodes}
\end{center}
\end{figure}
is a barplot of all FOIA exemption codes found in all of the emails.  As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made though affect more than 2,000 emails in the collection.   Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.   The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.  This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.  The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.

As with all other displays, the FOIA barplot will update in reaction to all filters making it a useful tool to identify how codes co-occur over any selection of emails.  Conversely, filtering on FOIA codes can be interesting to see who sees correspondence of varying sensitivities, especially B1 and B5.

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\caption{Top 20 terms appearing in all emails}
\label{fig:top20all}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. \TODO{Chris, check that I have described these two measures correctly}
The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.

The terms give some limited insight into the contents of the email.  As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.

As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is also challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).

\section{Filters}
\label{sect:Filters}
There are only three filters but each one affects all data displays.  As each filter is changed, every display redraws itself on the filtered data, immediately in reaction to the change.  In this way, the filters can be used together to focus on particular subsets of the emails or simply to observe how the display patterns change with the filter being applied.
\subsection{Time filtering}
A sliding time window 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{DateSliderImage}
\caption{Time slider: move either end, or move the interval}
\end{center}
\label{fig:timeSlider}
\end{figure}
filters the emails displayed to those whose date lies between the two end points.
The size of the time window is changed by  moving either end, either by mouse click and drag or by selecting an end to move with the arrow keys. The whole time window is moved by selecting the middle bar and either dragging it or moving it with the arrow keys.

This filter is simple but powerful.  All emails are displayed when the range covers all dates.  Moving the end points towards one another allows the user to focus the displays on any particular range, down to as fine as all emails on a single day.  With a fixed range of days, for example a two week period,  dragging the middle bar from left to right will have each display smoothly update over time.
%
%Time filtering within the app can be completed at the level of individual days. The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days. As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width. Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
\subsection{Sent or received}
A drop down menu 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{ToFromImage}
\caption{Filtering on whether Clinton sent or received the email, or either}
\label{fig:toFromFilter}
\end{center}
\end{figure}
is used to select those emails which Clinton sent, or received, or either sent or received.  The default is ``All Emails'', which is the same as no filter being applied for sent or received.  Choosing ``From Clinton'' selects for display only those mails sent by Secretary Clinton, choosing ``To  Clinton'' selects only those emails which she received.  This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported using a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.55\linewidth]{FOIABoxImage}
\caption{FOIA codes:  Emails containing any of these codes appear}
\label{fig:selectFOIA}
\end{center}
\end{figure}
Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.  Any email having one or more of these codes will appear in the displays.  To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list.  Or to have only the emails that  contain B1 redactions displayed, all other codes can deleted from the list.  Codes are returned to active display at any time by clicking in the display box and selecting them again from a drop down menu of the excluded codes.  Any number of codes  (one or more) can be selected to be included in the displays.

\subsection{Auxiliary information}
To demonstrate how other sources of information can easily be used to supplement the metadata found in the email archive, we consulted the State Department website and retrieved Secretary Clinton's official foreign travel schedule  \cite{ForeignSched}.  This information was added to the display as a simple check box ``Display Foreign Travel Schedule''.  When checked, light blue vertical bars appear on the  email volume time series display.  Appearing underneath the series, these mark time periods when Secretary Clinton was travelling outside the U.S. on State Department business.

\section{Some interactive analyses}
\label{sect:Analysis}
Each of the displays and filters described in the previous sections gives a slightly different and informative view of summary features of the email archive.  However, their analytic power is considerably amplified when used together.  Rather complex queries can be formed involving date, classification code, and whether Clinton is sending or receiving email.  Moreover, given a particular pattern in one display, its relation to that in another can be of considerable interest.   Together with access to the archive contents and an internet search engine, an enormous amount can be learned from these visual analytic tools.

In this section, we explore a few avenues of enquiry that might naturally arise.  At best this only hints at the analytic power that interactive filtering and display can bring to bear on an email archive.  To get a better appreciation, the user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}.  The tool is useful both for generating and for testing hypotheses.

\subsection{The last two years}
One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  This is done by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.  Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the resulting email volume time series over this period.  Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are relatively few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   Another series of the next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on this section of the emails.

Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:early2011B1}
\end{center}
\end{figure}
As can be seen, there continue to be highs and lows of email activity.   The highest peak now is at the far left.  This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.  Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing the Egyptian Crisis.

Slightly to the right of this peak, there are several more.   At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.  Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on this, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes. 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.  On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike is the rightmost spike of approximately 80 emails from Figure~{EmailVolumeB1JanApril2011}.   
Of course all other displays have been updating every time the time slider is adjusted.

Some sense of the content of the emails can be had from the updated term frequency displays shown in Figure~\ref{fig:TermFreqB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{TFIDFB1LibyaBuildup} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{TermFreqB1LibyaBuildup} 
\end{tabular}
\caption{Frequent terms: B1 only emails, Feb. 25 -- March 18, 2011}
\label{fig:TermFreqB1LibyaBuildup}
\end{center}
\end{figure}
Not surprisingly,  Libya and its leader figure prominently in this restricted set of emails.
